---
title: "How to Lose Less Money Playing the Lottery"
author: "Stephen Penrice"
date: "Friday, June 19, 2015"
output: ioslides_presentation
---


## Real Goal: Understanding players' preferences

- Players prefer some numbers, such as family birthdays, 7's, etc.
- Direct data exist but are not public, so how can we document this? 
- Can back out by relating number of winners to numbers drawn.
- Number of winners is only informative if total number of players is known. 
- Otherwise we can model parimutuel prizes as a function of the numbers drawn.

## Games Discussed Here

* Texas Cash 5  
    + Players and lottery choose 5 integers from 1 to 37
    + Target prize for 3 matches is $10.00
    + Jan. 1, 2012 to June 12, 2015
* Pennsylvania Cash 5
    + Players and lottery choose 5 integers from 1 to 43
    + Target prize for 3 matches is $11.50
    + Jan. 1, 2012 to June 12, 2015
* New Jersey Cash 5  
    + Players and lottery choose 5 integers from 1 to 43
    + Target prize for 3 matches is $15.00 
    + Sept. 14, 2014 to June 12, 2015
  

## Random vs. Actual, Texas

```{r}
tx.actual <- read.table("tx_data\\tx_cash5.csv",header=TRUE,sep=",") 
dhyper(3,5,32,5)
sum(tx.actual$winners_3)/sum(tx.actual$sales)
```

## Random vs. Actual, Texas

```{r}
draws <- nrow(tx.actual)
pop <- 150000
prob <- dhyper(3,5,32,5)
pool <- sum(tx.actual$prize_3*tx.actual$winners_3)/sum(tx.actual$sales)

set.seed(12345)
sim.winners <- rpois(draws,pop*prob)
sim.prizes <- round(pop*pool/sim.winners)
sim.pct.win <- sim.winners/pop

actual.prizes <- tx.actual$prize_3
actual.pct.win  <- tx.actual$winners_3/tx.actual$sales
```

## Random vs. Actual, Texas

```{r, echo=FALSE}
par(mfrow=c(2,2))
hist(sim.prizes,xlab="",main="Simulated Prize Distribution")
hist(actual.prizes,xlab="",main="Actual Prize Distribution")
hist(round(sim.pct.win,4),xlab="",main="Simulated Win Percentage Distribution",
     breaks = c(0.007,0.008,0.009,0.01,0.011,0.012,0.013,0.014,0.015,0.016,0.017))
hist(round(actual.pct.win,4),xlab="",main="Actual Win Percentage Distribution",
     breaks = c(0.007,0.008,0.009,0.01,0.011,0.012,0.013,0.014,0.015,0.016,0.017))
```


## Random vs. Actual, Pennsylvania

```{r}
pa.actual <- read.table("pa_data\\pa_cash5.csv",header=TRUE,sep=",") 

draws <- nrow(pa.actual)
pop <- 400000
prob <- dhyper(3,5,38,5)
pool <- prob*mean(pa.actual$prize_3)

set.seed(54321)
sim.winners <- rpois(draws,pop*prob)
sim.prizes <- 0.5*round(2*pop*pool/sim.winners)

actual.prizes <- pa.actual$prize_3
```

## Random vs. Actual, Pennsylvania

```{r,echo=FALSE}
par(mfrow=c(1,2))
hist(sim.prizes,xlab="",main="Simulated Prize Distribution", xaxt="n")
axis(side = 1, at = c(11.0,11.5,12.0) )
hist(actual.prizes,xlab="",main="Actual Prize Distribution")
```


## Modeling When Sales Data Are Available

Poisson Regression  

Model TX1:
$$ \ln(y) = \beta t + c +\ln(s)$$

where $y$ is the number of 3-match winners, $t$ is the sum of the numbers drawn, 
and $s$ is sales.   

Interpretation: $$\beta t + c$$ estimates the log of the percentage of players
who match 3 numbers.

## Modeling When Sales Data Are Available

```{r, results='asis'}
tx.actual$drawdate <- as.Date(as.character(tx.actual$drawdate))
tx.train <- tx.actual[tx.actual$drawdate <= '2014-06-12',]
tx.test <- tx.actual[tx.actual$drawdate > '2014-06-12',]
tx1 <- glm(winners_3~I(n1+n2+n3+n4+n5),data=tx.train,
              family=poisson,
              offset=log(sales))
knitr::kable(summary(tx1)$coefficients,digits=4)
```

## Modeling When Sales Data Are Available

Model TX2:
$$ \ln(y) = \beta_1 n_1 + \beta_2 n_2 + \beta_3 n_3 + \beta_4 n_4 + \beta_5 n_5 + c +\ln(s)$$

where $y$ is the number of winners, $t$ is the sum of the numbers drawn, and $s$ is 
sales.

## Modeling When Sales Data Are Available

```{r, results='asis',echo=FALSE}
tx2 <- glm(winners_3~n1+n2+n3+n4+n5,data=tx.train,
              family=poisson,
              offset=log(sales))
knitr::kable(summary(tx2)$coefficients,digits=4)
```

## Modeling Prize Amounts

Without sales data, we resort to modeling prizes as a function of the numbers
drawn. Linear models are not justifiable from an inference point of view, but
they could provide good predictions.

Model PA1:  
$$ y = \beta t + c$$  
where $y$ is the 3-match prize and $t$ is the sum of the numbers drawn.

## Modeling Prize Amounts
```{r}
pa.actual$drawdate <- as.Date(as.character(pa.actual$drawdate))
pa.train <- pa.actual[pa.actual$drawdate <= '2014-06-12',]
pa.test <- pa.actual[pa.actual$drawdate > '2014-06-12',]
```

```{r, results='asis'}
pa1 <- lm(prize_3~I(n1+n2+n3+n4+n5),data=pa.train)
knitr::kable(summary(pa1)$coefficients,digits=4)
print(summary(pa1)$r.squared)
```

## Some Refinements

Besides small numbers, players tend to choose combinations of numbers that are
physically close on a playslip. So we'll add variables to proxy combinations
selected along a row or column.

```{r,warning=FALSE}
library(dplyr,warn.conflicts=FALSE)
library(magrittr)

pa.train <- mutate(pa.train,range=n5-n1) %>%
           mutate(spacing = 
                  sqrt((((n2-n1)^2+(n3-n2)^2+(n4-n3)^2+(n5-n4)^2)/4)-
                               ((n5-n1)/4)^2)) 
```


## Some Refinements

Model PA2:  
$$ y = \beta_1 t + \beta_2 r + \beta_3 s + c$$  
where $y$ is the 3-match prize, $t$ is the sum of the numbers drawn, $r$ is
the range of the numbers drawn, and $s$ is the standard deviation of the gaps
between consecutive numbers drawn.

## Some Refinements

```{r,results='asis'}
pa2 <- lm(prize_3~I(n1+n2+n3+n4+n5)+range+spacing,data=pa.train)
knitr::kable(summary(pa2)$coefficients,digits=4)
print(summary(pa2)$r.squared)
```


## Validating on Recent Data

## Validating on New Jersey Data

## Comment on Deriving Player Choices

## Getting Data

## References

- My github repo   
<https://github.com/lotterdata/mis_presentation>  

- Number selection algorithm exploiting this idea  
<http://www.r-bloggers.com/picking-lotto-numbers/> 

- A recent article in a similar vein  
<http://www.r-bloggers.com/simple-data-science-to-maximize-return-on-lottery-investment/>
